---
title: "Recreating N. MacLeod's CVA example, More Matrix based"
output: html_notebook
---

# About

Norm MacLeod's [PaleoMath article on CVA](https://www.palass.org/sites/default/files/media/palaeomath_101/article_11/article_11.pdf) includes a nice simplified example where he does CVA on a reduced version of the iris data set, treating the CVA as a two-step rotation. Here I recreate this analysis in R.

# Libraries

```{r}
library(MASS)
library(tidyverse)
library(magrittr)
```

# Analysis

First generate the reduced iris data set -- only "Sepal" variables and just 25 specimens per species. NOTE: in the MacLeod article his figures read "Petal Length" and "Petal Width" but comparison of the data suggest he likely mislabeled the plots and the data is "Sepal Length" and "Sepal Width"


```{r}
# Take Sepal data for first 25 specimens of each species
reduced.iris <-
  iris %>%
  select(-starts_with("Petal")) %>%  # drop petal variables
  group_by(Species) %>%
  do(head(., 25)) %>%
  ungroup

species <- reduced.iris$Species
data <- reduced.iris %>% select(-Species) %>% as.matrix

# center data
data <- scale(data, center=TRUE, scale=FALSE)
```
  
Plot data

```{r}
variate.plot <-
  data %>% as.data.frame %>% mutate(Species = species) %>%
  ggplot() +
  geom_point(aes(x = Sepal.Length, y = Sepal.Width, color=Species)) +
  coord_fixed()

variate.plot
```


```{r}
first.PCA <- prcomp(data)

first.PCA.plot <- 
  first.PCA$x %>% as.data.frame %>% mutate(Species = species) %>%
  ggplot() +
  geom_point(aes(x = PC1, y = PC2, color=Species)) +
  coord_fixed()

first.PCA.plot
```

```{r}

scaled.scores <- first.PCA$x %*% diag(1/first.PCA$sdev)
colnames(scaled.scores) <- c("scaled.PC1", "scaled.PC2")

first.PCA.scaled.scores <- 
  scaled.scores %>% as.data.frame %>%
  mutate(Species = species)

first.PCA.group.means <-
  first.PCA.scaled.scores %>%
  group_by(Species) %>%
  summarize_all(mean)

scaled.PCA.plot <-
  ggplot() +
  geom_point(aes(x=scaled.PC1, y = scaled.PC2, color=Species),
             data = first.PCA.scaled.scores) + 
  geom_point(aes(x=scaled.PC1, y = scaled.PC2, color=Species),
             data = first.PCA.group.means, size=4, alpha=0.5) +
  coord_fixed()

scaled.PCA.plot  
```

```{r}
second.PCA <- prcomp(first.PCA.group.means %>% select(-Species))

pts <- matrix(c(1,0,-1,0,0,1,0,-1),byrow = TRUE, nrow=4)
second.PCA.axes <- as.data.frame(pts %*% second.PCA$rotation)


ggplot(first.PCA.group.means, aes(x=scaled.PC1, y=scaled.PC2)) +
  geom_point(aes(shape=Species, color=Species), size=4, alpha=0.8) + 
  geom_segment(aes(x=0,y=0,xend=PC1, yend=PC2),data=second.PCA.axes, inherit.aes = FALSE) +
  coord_fixed()
```


```{r}

CVA.scores <- scaled.scores %*% second.PCA$rotation
colnames(CVA.scores) <- c("CV1", "CV2")


CVA.scores %>% as.data.frame %>% mutate(Species = species) %>%
  ggplot() + 
  geom_point(aes(x=CV1, y = CV2, color=Species))

```



```{r}
ggplot() + 
  geom_point(aes(x=scaled.PC1, y=scaled.PC2, color=Species),
             data=first.PCA.scaled.scores) +
  geom_point(aes(x=scaled.PC1, y=scaled.PC2, color=Species), 
             data=first.PCA.group.means, size=4, alpha=0.4) +
  geom_segment(aes(x=0,y=0,xend=PC1, yend=PC2),data=second.PCA.axes,inherit.aes = FALSE) +
  coord_fixed()
```

```{r}
unscaled.CVA.axes <- 
  as.matrix(second.PCA.axes) %*% diag(first.PCA$sdev) 


unscaled.CVA.df <-
  unscaled.CVA.axes %>%
  as.data.frame

names(unscaled.CVA.df) <- c("CV1", "CV2")


first.PCA.plot+ 
  geom_segment(aes(x=0,y=0,xend=CV1, yend=CV2),data=unscaled.CVA.df,inherit.aes = FALSE)
```

```{r}
CVA.in.variate.axes <- unscaled.CVA.axes %*% solve(pca.I$rotation)

CVA.in.variate.df <-
  CVA.in.variate.axes %>%
  as.data.frame

names(CVA.in.variate.df) <- c("CV1", "CV2")

variate.plot + 
  geom_segment(aes(x=0,y=0,xend=CV1, yend=CV2),
               data=CVA.in.variate.df,
               inherit.aes = FALSE) +
  coord_fixed()


```

