---
title: "Introduction to dplyr"
author: "Paul M. Magwene"
output: 
  beamer_presentation:
    colortheme: "beaver"
    fonttheme: "structurebold"
fontsize: "9pt"    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, comment = "")
options(width=60)
```

```{r}
library(tidyverse)
```

## What is `dplyr`?

`dplyr` is a package that provides a "grammar for data manipulation"

\bigskip

Key "verbs" in the `dplyr` package:

  * `select()`
  * `filter()`
  * `mutate()`
  * `arrange()`
  * `summarize()`
  * `group_by()`
  
\bigskip

All these functions return new data frames instead of modifying existing data frames

## `select()` subsets columns

```{r, echo=TRUE}
names(iris)
```


```{r, echo = TRUE}
# select two columns
select(iris, Sepal.Length, Petal.Length) %>% head(3)
```


```{r, echo = TRUE}
# select everything BUT the species column
select(iris, -Species) %>% head(3)
```

## `select()` has some specialized functions for powerful filtering

```{r, echo = TRUE}
select(iris, starts_with("Petal")) %>% head(3)
```


```{r, echo = TRUE}
select(iris, ends_with("Length")) %>% head(3)
```



## `filter()` selects rows that match criteria

```{r, echo = TRUE}
# get only the I. setosa specimens
filter(iris, Species == "setosa") %>% head(3)
```


```{r}
# filter on mulitple criteria 
filter(iris, Species == "setosa", Sepal.Length < 5)
```


## `mutate()` 
