---
title: "Simulating Sampling Distributions "
author: "Paul M. Magwene"
output: 
  beamer_presentation:
    colortheme: "beaver"
    fonttheme: "structurebold"
fontsize: "10pt"    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE)
options(width=50)
```



## Population of interest

```{r}
library(tidyverse)
set.seed(20171114)

height.data <- read_csv("https://github.com/bio304-class/bio304-fall2017/raw/master/datasets/simulated-US-male-height.csv")

ggplot(height.data, aes(x = height)) +
  geom_histogram(aes(y = ..density..), bins=50, alpha=0.5) + 
  geom_density() +
  labs(x = "Height (cm)", title = "Distribution of Heights in the Population of Interest")
```


## A random sample of size 25

```{r}
true.values <- 
  height.data %>% 
  summarize(mean.height = mean(height), 
            sd.height = sd(height))

sample.a <-
  height.data %>%
  sample_n(25)

sample.a %>%
  ggplot(aes(x = height)) + 
  geom_histogram(data=height.data, aes(x = height, y = ..density..), alpha=0.25, bins=50) + 
  geom_histogram(aes(y = ..density..), fill = 'steelblue', alpha=0.75, bins=9) + 
  geom_vline(xintercept = true.values$mean.height, linetype = "solid") + 
  geom_vline(xintercept = mean(sample.a$height), linetype = "dashed") + 
  labs(x = "Height (cm)", y = "Density", 
       title = "Distribution of heights in the underlying population (grey) and for a single sample of size 25 (blue).\nThe true population mean is shown as a solid line, the sample mean is shown as a dashed line")
```

## A second random sample of size 25

```{r}
true.values <- 
  height.data %>% 
  summarize(mean.height = mean(height), 
            sd.height = sd(height))

sample.b <-
  height.data %>%
  sample_n(25)

sample.b %>%
  ggplot(aes(x = height)) + 
  geom_histogram(data=height.data, aes(x = height, y = ..density..), alpha=0.25, bins=50) + 
  geom_histogram(aes(y = ..density..), fill = 'steelblue', alpha=0.75, bins=9) + 
  geom_vline(xintercept = true.values$mean.height, linetype = "solid") + 
  geom_vline(xintercept = mean(sample.b$height), linetype = "dashed") + 
  labs(x = "Height (cm)", y = "Density", 
       title = "Distribution of heights in the underlying population (grey) and for a single sample of size 25 (blue).\nThe true population mean is shown as a solid line, the sample mean is shown as a dashed line")
```

## A third random sample of size 25

```{r}
true.values <- 
  height.data %>% 
  summarize(mean.height = mean(height), 
            sd.height = sd(height))

sample.c <-
  height.data %>%
  sample_n(25)

sample.c %>%
  ggplot(aes(x = height)) + 
  geom_histogram(data=height.data, aes(x = height, y = ..density..), alpha=0.25, bins=50) + 
  geom_histogram(aes(y = ..density..), fill = 'steelblue', alpha=0.75, bins=9) + 
  geom_vline(xintercept = true.values$mean.height, linetype = "solid") + 
  geom_vline(xintercept = mean(sample.c$height), linetype = "dashed") + 
  labs(x = "Height (cm)", y = "Density", 
       title = "Distribution of heights in the underlying population (grey) and for a single sample of size 25 (blue).\nThe true population mean is shown as a solid line, the sample mean is shown as a dashed line")
```

## Sampling distribution of the mean: Distribution of sample means across many random samples


```{r}
rsample.stats <- function(x, n, var.name) {
  sample_x <- sample_n(x, n)
  data_frame(sample.size = n, 
             sample.mean = mean(sample_x[[var.name]]), 
             sample.sd = sd(sample_x[[var.name]]))
}

df.samples.of.25 <-
  rerun(500,  rsample.stats(height.data, 25, "height")) %>%
  bind_rows()

se.mean.table <-
  df.samples.of.25 %>%
  summarize(mean.of.means = mean(sample.mean),
            sd.of.means = sd(sample.mean))

ggplot(df.samples.of.25, aes(x = sample.mean, y = ..density..)) +
  geom_histogram(bins=25, fill = 'firebrick', alpha=0.5) + 
  geom_vline(xintercept = true.values$mean.height, linetype = "dashed") + 
  geom_vline(xintercept = se.mean.table$mean.of.means + se.mean.table$sd.of.means,
             linetype = 'dashed', color = 'red') +
  geom_vline(xintercept = se.mean.table$mean.of.means - se.mean.table$sd.of.means,
             linetype = 'dashed', color = 'red') +  
  labs(x = "Sample means", y  = "Density",
       title = "Distribution of mean heights for 500 samples of size 25")

```



## Comparing the population distribution to the sampling distribution of the mean


```{r}
ggplot(df.samples.of.25, aes(x = sample.mean, y = ..density..)) +
  geom_histogram(data=height.data, aes(x = height, y = ..density..), alpha=0.25, bins=50) + 
  geom_histogram(bins=50, fill = 'firebrick', alpha=0.5) + 
  geom_vline(xintercept = true.values$mean.height, linetype = "dashed") + 
  geom_vline(xintercept = se.mean.table$mean.of.means + se.mean.table$sd.of.means,
             linetype = 'dashed', color = 'red') +
  geom_vline(xintercept = se.mean.table$mean.of.means - se.mean.table$sd.of.means,
             linetype = 'dashed', color = 'red') +    
  labs(x = "Heights (gray) or Sample mean heights (red)", y  = "Density",
       title = "Comparison of the population distribution of heights and\nthe sampling distribution of mean heights for samples of size 25")


```


## Sampling Distributions and Standard Errors


### Sampling Distributions

The **sampling distribution of a statistic of interest** is the probability distribution of a given statistic for samples of a given size. 

### Standard Errors

The **standard error of a statistic of interest** is the standard deviation of the sampling distribution for the given statistic

### Uncertainty

Together, the sampling distribution and associated standard error for a given statistic are key measures of the uncertainty in statistical estimates

